---
title: "Downloading data with vegbankr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading data with vegbankr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction to `vegbankr`

This package is an R client for VegBank, the vegetation plot database of
the Ecological Society of America's [Panel on Vegetation
Classification](https://esa.org/vegpanel/), hosted by the [National
Center for Ecological Analysis and
Synthesis](https://www.nceas.ucsb.edu) (NCEAS). VegBank contains
vegetation plot data, community types recognized by the U.S. National
Vegetation Classification and others, and all ITIS/USDA plant taxa along
with other taxa recorded in plot records.

As a VegBank API client, the `vegbankr` package currently supports querying and downloading
vegetation plot records, in addition to validating and uploading new
data to the VegBank database.

Many of the functions utilize the following arguments:

| Argument | Description |
|---|---|
| `vb_code` | A VegBank code. For example, an observation code (`ob.*`) or project code (`pj.*`) |
| `detail` | Level of detail returned: `"full"` includes all available fields; the default returns a summary subset |
| `with_nested` | If `TRUE`, nested child records (e.g. taxon observations, stratum data) are included as list columns |
| `limit` | Maximum number of records to return (default: 100) |
| `offset` | Number of records to skip before returning results; useful for pagination |
| `sort` | Field name to sort by; prefix with `-` for descending order (e.g. `"-obs_count"`) |
| `search` | Optional search string for filtering results | 


```{r, message = FALSE, echo = FALSE}
library(vegbankr)
```

## Plot Observation Data

The `vb_get_plot_observations()` function allows users to download
vegetation plot data from VegBank. 

### Single Plot Observation by Plot Code

To retrieve a specific plot observation record using its "ob" code, use
the following command:

```{r}
# Retrieve a specific plot observation
ob.135454 <- vb_get_plot_observations(vb_code = "ob.135454", detail = "full",
  with_nested = TRUE)

# Preview the downloaded data
head(ob.135454)
```

Omitting `detail = "full"` or setting `with_nested = FALSE` returns a
smaller summary better for quick browsing or larger downloads.

### Plot Observations for a Project

You can also download multiple plot observations for a specific project.
For example, to retrieve 100 plot observations from the Southwest GAP,
Nevada Project (pj.10510):

```{r}
# Retrieve all plot observations for a specific project
pj.10510 <- vb_get_plot_observations(vb_code = "pj.10510")

# Preview the data
head(pj.10510)

```

## Project Data

`vb_get_projects()` returns information about projects established to collect vegetation plot data.

### Search by Project Name

This example retrieves all projects whose name contains "GAP", sorted in descending order by observation count so that the most data-rich projects appear first.

```{r}
vb_get_projects(search = "GAP", sort = "-obs_count")
```

### Plot Observations by Project Code

Once you have identified a project code, you can pass it directly to `vb_get_plot_observations`. This example retrieves the first 100 records of plot observations associated
with the project `pj.11044` (Pennsylvania HPD Delaware Water Gap), sorted by the author's observation code.

```{r}
obs <- vb_get_plot_observations("pj.11044", sort = "author_obs_code")
obs
```

## Taxon Observations

`vb_get_taxon_observations()` retrieves the individual plant taxon records associated with a given plot observation. Each row represents one taxon recorded in the plot

### Taxon Observations by Plot

This example retrieves the taxon (plant) observations associated with the plot `ob.135454`

```{r}
vb_get_taxon_observations("ob.135454")
```

## Plant Community Concepts

The example below searches for community concepts that include the genus
*Sequoiadendron*.

```{r}

vb_get_community_concepts(search = "sequoiadendron")

```

## Other Function Options

### Changing Limit & Offset

To download more than 100 records, increase the `limit` argument. To page through a large results in set chunks, combine `limit` and `offset`

```{r}

# Download up to 500 records
vb_get_plot_observations(vb_code = "pj.10510", limit = 500)

# Download the second page of 100 records
vb_get_plot_observations(vb_code = "pj.10510", limit = 100, offset = 100)

```


### Sorting Results

### Manipulating Downloaded Data with `dplyr`

Because downloaded VegBank data is saved as dataframes, the data can be
manipulated using base R or dplyr functions. Below we highlight a few
possible data manipulations.

```{r, message = FALSE}
library(dplyr)
```

#### ! Need to think of exampls !