---
title: "Downloading data with vegbankr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Downloading data with vegbankr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

## Introduction to `vegbankr`

This package is an R client for VegBank, the vegetation plot database of
the Ecological Society of America's [Panel on Vegetation
Classification](https://esa.org/vegpanel/), hosted by the [National
Center for Ecological Analysis and
Synthesis](https://www.nceas.ucsb.edu) (NCEAS). VegBank contains
vegetation plot data, community types recognized by the U.S. National
Vegetation Classification and others, and all ITIS/USDA plant taxa along
with other taxa recorded in plot records. As a VegBank API client, the
`vegbankr` package currently supports querying and downloading
vegetation plot records and other supporting information from the
VegBank database, and will soon support validating and uploading new
data to the VegBank database as well.

```{r, message = FALSE, echo = FALSE}
library(vegbankr)
```

### Downloading Plot Observation Data with `vb_get_observations()`

#### Download Single Plot Observation by Plot Code

Grab a single plot observation record based on its "ob" code.

```{r}
ob.135454 <- vb_get_plot_observations(vb_code = "ob.135454", detail = "full",
  with_nested = TRUE)
head(ob.135454)
```

#### Download Plot Observations for a Project with `vb_get_plot_observations()`

This example downloads 100 plot observations for the Southwest GAP,
Nevada Project (pj.10510).

```{r}
vb_get_plot_observations(vb_code = "pj.10510")

# To download more than 100 records, increase the limit argument
# vb_get_plot_observations(vb_code = "pj.10510", limit = 500)
```

### Downloading Project Data with `vb_get_projects()`

#### Search by Project Name

This example searches for "GAP" related projects, and returns them in
descending order by observation count.

```{r}
vb_get_projects(search = "GAP", sort = "-obs_count") |>
  dplyr::select(pj_code, project_name, obs_count)
```

#### Search by Project Code

This example gets the first 100 records of plot observations associated
with the project `pj.11044` (Pennsylvania HPD Delaware Water Gap). It is
then sorted by the author_obs_code

```{r}
obs <- vb_get_plot_observations("pj.11044", sort = "author_obs_code",
  limit = 100)
obs |> dplyr::count(state_province)
```

### Downloading Data by Taxon with `vb_get_taxon_observations()`

#### Get Taxon by Plot Observation

This example gets the taxon (plant) observations associated with a plot 
observation (ob.135454), and then displays the observations in order based on the plant code of the
current taxon interpretation.

```{r}
vb_get_taxon_observations("ob.135454") |>
  dplyr::arrange(int_curr_plant_code) |>
  print(n = 35)
```

#### Get Taxon by Name

This example searches for community concepts with the string "*sequoiadendron*".

```{r}

sequoia_communities <- vb_get_community_concepts(search = "sequoiadendron")

```
