---
title: "Uploading plot data with vegbankr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Uploading data with vegbankr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, message = FALSE} 
library(dplyr)
library(DT)

docs <- read.csv("../inst/loader-table-fields.csv")

docs$required <- factor(docs$required, levels = c("required", "best practice", "commonly used", "sometimes used"))

docs <- docs %>% 
  arrange(required)
```

## Introduction to `vegbankr`

This package is an R client for VegBank, the vegetation plot database of
the Ecological Society of America's [Panel on Vegetation
Classification](https://esa.org/vegpanel/), hosted by the [National
Center for Ecological Analysis and
Synthesis](https://www.nceas.ucsb.edu) (NCEAS). VegBank contains
vegetation plot data, community types recognized by the U.S. National
Vegetation Classification and others, and all ITIS/USDA plant taxa along
with other taxa recorded in plot records. As a VegBank API client, the
`vegbankr` package currently supports querying and downloading
vegetation plot records and other supporting information from the
VegBank database, and will soon support validating and uploading new
data to the VegBank database as well.

To use `vegbankr` to upload data, there are 3 key steps:

1.  Model and transform your data to the vegbank loader table format
2.  Validate your data
3.  Upload your data using `vb_upload_plot_observations(...)`

This vignette will walk through these 3 steps, with an emphasis on
modeling and validating data.

## Vegbank Loader Tables

Loader tables are the data format that is used to upload data into
VegBank. In order to publish your data to VegBank, the first step is to
model whatever format your data is in to the loader table format, and
then transform the data into that format. Modeling your data means
identifying how each piece of information in your original dataset (like
species names, plot locations, survey dates) corresponds to specific
fields in the VegBank loader table format. There are a total of 12
loader tables that can be used for plot observation data, though not all
are required for data ingest. In this section, the loader tables and
their fields will be described, in the order in which it is recommended
to prepare them.

Each loader table you create is an R `data.frame` that is eventually
passed to a VegBank upload function. In the documentation below,
interactive tables display the allowed `field` (column names), whether
the field is required, best practice, commonly used, or sometimes used,
and a description of the field.

There are number of fields that act as codes which are used primary and secondary keys to link loader tables together. Codes that begin with `user_` are supplied by the data uploader. Codes that begin with `vb_` are created by the database upon data upload. 

### Projects

This table stores information about a project established to collect
vegetation plot data. The `user_pj_code` is the project code primary
key, and is found as a foreign key in several other tables. An example
project code might be `MOJA` with project name "Mojave Desert Vegetation
Surveys."

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "projects") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Parties
 
 The Parties loader table is used to upload new parties (people) associated with plots, projects, taxa, and classifications. The primary key is `user_py_code` which is used as a foreign key in the Contributors loader table. Once uploaded, VegBank will create a `vb_py_code` for each party to be used in the Contributors table.

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "parties") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Contributors

The contributors loader table is fairly code heavy, but is closely linked to
both the Projects table and the Parties table, and is used to link parties (people) with their contributions to plots, projects, taxa, and classifications.

`user_py_code` is a foreign key to the `parties` table, so any values present in this field must be present there as well. Optionally, instead of `user_py_code`, `vb_py_code` can be used if the party is already in vegbank. One of `user_py_code` or `vb_py_code` must be present, but having both in the same row is disallowed.

`vb_ar_code` is the vegbank role code - a code in the format `ar.{nn}`. A table of allowed values and their meanings is listed below the loader table variables.

Finally, the `contributor_type` indicates whether this contributor is linked to an Observation, Project, or Classification. The `record_identifier` value will depend on which of these three types the contributor is associated with. If the contributor should be associated with a project with `user_pj_code` `MOJA`, the `record_identifier` for that contributor should be `MOJA` and the contributor type should be Project. This transitively will also associate the contributor with all plots associated with that project. If the contributor should only be associated with a particular observation with `user_ob_code` `MOJA_0214`, the record identifier is that observation identifier, and the `contributor_type` should be Observation.


```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "contributors") %>% 
  select(-loader_table) %>% 
  datatable()
```

```{r, echo = FALSE}

roles <- dplyr::tribble(
  ~ar_code, ~role_name,
  "ar.16",  "Author",
  "ar.17",  "Contact",
  "ar.18",  "PI",
  "ar.19",  "Data Manager",
  "ar.34",  "Classifier",
  "ar.36",  "Plot author",
  "ar.38",  "Co-PI",
  "ar.39",  "Computer (automated)",
  "ar.40",  "Consultant",
  "ar.43",  "Field assistant",
  "ar.44",  "Guide",
  "ar.45",  "Land owner",
  "ar.46",  "Not specified",
  "ar.47",  "Not specified/Unknown",
  "ar.48",  "Passive observer",
  "ar.50",  "Plot contributor",
  "ar.51",  "Publication author",
  "ar.53",  "Research advisor",
  "ar.54",  "System manager",
  "ar.55",  "Taxonomist",
  "ar.56",  "Data aggregator"
) 

datatable(roles)

```

### Plot Observations

The plot observations loader table contains all data that is consistent across a plot. This includes information on the plot name, location, physical features, non-vegetation cover, etc. This table has many optional fields that may or may not be applicable to your project.

Similar to the pattern described in contributors, one of `vb_pl_code` or `user_pl_code` is required, and only one of those two fields may be used for each row. `vb_pl_code` would only be used if the intention is to add a new observation record to the same plot. These codes are used as foreign keys in various other tables. `author_plot_code` is often the same as these codes (eg: `MOJA_0214`) but there they could be different for valid reasons. `author_plot_code` is prominently displayed in the user VegBank interface as the plot identifier. 

`user_ob_code` is the primary key for an observation on a plot, and may be the same as the plot code if there is only one observation of each plot. If there are multiple observations on the same plot, however, `user_ob_code` must be unique for each observation. 

`user/vb_pj_code` can link a plot and it's observation back to a project. Values in these fields must be present either in the `projects` loader table (`user_pj_code`) or VegBank if the project is already uploaded (`vb_pj_code`). 

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "plot_observations") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Community Classifications

The community classifications loader table contains the community classification of an observation. 

The primary key is `user_cl_code`. The foreign key `user_ob_code` corresponds to the key present in the plot observations loader table, and is required. All values in this field must also be present in the plot observations table.

`vb_cc_code` is the VegBank community concept identifier, a required field, the value of which must already be present in VegBank. To retrieve a list of possible `vb_cc_code` values, use the `vegbankr` function `vb_get_community_concepts`.

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "community_classifications") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Strata Cover

This loader table contains data from a plot observation of the plant names, cover, and strata in a given plot. The `user_ob_code` is a required foreign key that links the plant to a plot observation. All values in this field must be present in the plot observations loader table. `user_tm_code` is a key that is unique for each combination of `user_ob_code`, plant name, and strata in this table. `user_to_code` is a key that is unique for each combination of `user_ob_code` and plant name - so `user_to_code` may be repeated in this table if a plant exists in multiple strata in a plot observation.

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "strata_cover_data") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Strata

### Taxon Interpretations

### Distrubances

### Soils

The Soils loader table is used to describe soils collected from a plot. This includes information on soil horizons, texture, and chemical properties.

```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "soils") %>% 
  select(-loader_table) %>% 
  datatable()
```

### Stem Data

The Stem Data loader table is used to describe individual plant stems measured at a plot. 
```{r, echo = FALSE}
docs %>% 
  filter(loader_table == "stem_data") %>% 
  select(-loader_table) %>% 
  datatable()
```


### References

